# (PART) Manipulations {-}

# Basic String Manipulations {#manip}

## Introduction

In this chapter you will learn about the different functions to do what I call
_basic manipulations_. By "basic" I mean transforming and processing strings
in such way that do not require the use of [regular expressions](#intro). More 
advanced manipulations involve defining patterns of text and matching such
patterns. This is the essential idea behind regular expressions, which is 
the content of part 5 in this book.


## Basic Manipulations

Besides creating and printing strings, there are a number of very handy 
functions in R for doing some basic manipulation of strings. In this chapter    
we will review the following functions:

| Function       | Description                      |
|:---------------|:---------------------------------|
| `nchar()`      | number of characters             |
| `tolower()`    | convert to lower case            |
| `toupper()`    | convert to upper case            |
| `casefold()`   | case folding                     |
| `chartr()`     | character translation            |
| `abbreviate()` | abbreviation                     |
| `substring()`  | substrings of a character vector |
| `substr()`     | substrings of a character vector |

 
## Counting characters

One of the main functions for manipulating character strings is `nchar()` 
which counts the number of characters in a string. In other words, `nchar()` 
provides the length of a string:

```{r nchar_ex1}
# how many characters?
nchar(c("How", "many", "characters?"))

# how many characters?
nchar("How many characters?")
```

Notice that the white spaces between words in the second example are also 
counted as characters. 

It is important not to confuse `nchar()` with `length()`. While the former 
gives us the __number of characters__, the later only gives the number of 
elements in a vector.

```{r length_ex1}
# how many elements?
length(c("How", "many", "characters?"))

# how many elements?
length("How many characters?")
```


## Casefolding

R comes with three functions for text casefolding. 

1. `tolower()`
2. `toupper()`
3. `casefold()`

The function `tolower()` converts any upper case characters into lower case:

```{r tolower_ex}
# to lower case
tolower(c("aLL ChaRacterS in LoweR caSe", "ABCDE"))
```

The opposite function of `tolower()` is `toupper`. As you may guess, this 
function converts any lower case characters into upper case:

```{r toupper_ex}
# to upper case
toupper(c("All ChaRacterS in Upper Case", "abcde"))
```

The third function for case-folding is `casefold()` which is a wrapper for both `tolower()` and `toupper()`. Its uasge has the following form:

```
casefold(x, upper = FALSE)
```

By default, `casefold()` converts all characters to lower case, but you can 
use the argument `upper = TRUE` to indicate the opposite (characters in upper 
case):

```{r casefold_ex}
# lower case folding
casefold("aLL ChaRacterS in LoweR caSe")

# upper case folding
casefold("All ChaRacterS in Upper Case", upper = TRUE)
```

I've found the case-folding functions to be very helpful when I write functions
that take a character input which may be specified in lower or upper case, or 
perhaps as a mix of both cases. For instance, consider the function 
`temp_convert()` that takes a temperature value in Fahrenheit degress, and
a character string indicating the name of the scale to be converted.

```{r}
temp_convert <- function(deg = 1, to = "celsius") {
  switch(to,
         "celsius" = (deg - 32) * (5/9),
         "kelvin" = (deg + 459.67) * (5/9),
         "reaumur" = (deg - 32) * (4/9),
         "rankine" = deg + 459.67)
}
```

Here is how you call `temp_convert()` to convert 10 Fahrenheit degrees into 
celsius degrees:

```{r}
temp_convert(deg = 10, to = "celsius")
```

`temp_convert()` works fine when the argument `to = 'celsius'`. But what 
happens if you try `temp_convert(30, 'Celsius')` or 
`temp_convert(30, 'CELSIUS')`?

To have a more flexible function `temp_convert()` you can apply `tolower()`
to the argument `to`, and in this way guarantee that the provided string by
the user is always in lower case:

```{r}
temp_convert <- function(deg = 1, to = "celsius") {
  switch(tolower(to),
         "celsius" = (deg - 32) * (5/9),
         "kelvin" = (deg + 459.67) * (5/9),
         "reaumur" = (deg - 32) * (4/9),
         "rankine" = deg + 459.67)
}
```

Now all the following three calls are equivalent:

```{r, eval=FALSE}
temp_convert(30, 'celsius')
temp_convert(30, 'Celsius')
temp_convert(30, 'CELSIUS')
```



## Translating characters

There's also the function `chartr()` which stands for _character translation_. `chartr()` takes three arguments: an `old` string, a `new` string, and a 
character vector `x`:

```
chartr(old, new, x)
```

The way `chartr()` works is by replacing the characters in `old` that appear 
in `x` by those indicated in `new`. For example, suppose we want to translate 
the letter `"a"` (lower case) with `"A"` (upper case) in the sentence 
`"This is a boring string"`:

```{r chartr_ex1}
# replace 'a' by 'A'
chartr("a", "A", "This is a boring string")
```

It is important to note that `old` and `new` must have the same number of 
characters, otherwise you will get a nasty error message like this one:

```{r chartr_ex2, error = TRUE}
# incorrect use
chartr("ai", "X", "This is a bad example")
```

Here's a more interesting example with `old = "aei"` and `new = "\#!?"`. 
This implies that any `'a'` in `'x'` will be replaced by `'\#'`, any `'e'` in 
`'x'` will be replaced by `'?'`, and any `'i'` in `'x'` will be replaced by 
`'?'`:

```{r chartr_ex3}
# multiple replacements
crazy <- c("Here's to the crazy ones", "The misfits", "The rebels")
chartr("aei", "#!?", crazy)
```


## Abbreviating strings

Another useful function for basic manipulation of character strings is 
`abbreviate()`. Its usage has the following structure:

```
abbreviate(names.org, minlength = 4, dot = FALSE, strict = FALSE,
            method = c("left.keep", "both.sides"))
```

Although there are several arguments, the main parameter is the character 
vector (`names.org`) which will contain the names that we want to abbreviate:

```{r abbreviate_ex}
# some color names
some_colors <- colors()[1:4]
some_colors

# abbreviate (default usage)
colors1 <- abbreviate(some_colors)
colors1

# abbreviate with 'minlength'
colors2 <- abbreviate(some_colors, minlength = 5)
colors2

# abbreviate
colors3 <- abbreviate(some_colors, minlength = 3, method = "both.sides")
colors3
```


A common use for `abbreviate()` is when plotting names of objects or variables
in a graphic. I will use the built-in data set `mtcars` to show you a simple
example with a scatterplot between variables `mpg` and `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, type = "n")
text(mtcars$mpg, mtcars$disp, rownames(mtcars))
```

The names of the cars are all over the plot. In this situation you may want to
consider using `abbreviate()` to shrink the names of the cars and produce a 
less "crowded" plot:

```{r}
plot(mtcars$mpg, mtcars$disp, type = "n")
text(mtcars$mpg, mtcars$disp, abbreviate(rownames(mtcars)))
```


## Replacing strings

One common operation when working with strings is the extraction and 
replacement of some characters. There a various ways in which characters can
be replaced. If the replacement is based on the positions that characters 
occupy in the string, you can use the functions `substr()` and `substring()`

`substr()` extracts or replaces substrings in a character vector. Its usage has 
the following form:

```
substr(x, start, stop)
```

`x` is a character vector, `start` indicates the first element to be replaced, 
and `stop` indicates the last element to be replaced:

```{r substr_ex}
# extract 'bcd'
substr("abcdef", 2, 4)

# replace 2nd letter with hash symbol
x <- c("may", "the", "force", "be", "with", "you")
substr(x, 2, 2) <- "#"
x

# replace 2nd and 3rd letters with happy face
y = c("may", "the", "force", "be", "with", "you")
substr(y, 2, 3) <- ":)"
y

# replacement with recycling
z <- c("may", "the", "force", "be", "with", "you")
substr(z, 2, 3) <- c("#", "```")
z
```


Closely related to `substr()` is the function `substring()` which extracts or 
replaces substrings in a character vector. Its usage has the following form:

```
substring(text, first, last = 1000000L)
```

`text` is a character vector, `first` indicates the first element to be 
replaced, and `last` indicates the last element to be replaced:

```{r substring_ex}
# same as 'substr'
substring("ABCDEF", 2, 4)
substr("ABCDEF", 2, 4)

# extract each letter
substring("ABCDEF", 1:6, 1:6)

# multiple replacement with recycling
text6 <- c("more", "emotions", "are", "better", "than", "less")
substring(text6, 1:3) <- c(" ", "zzz")
text6
```



## Set Operations

R has dedicated functions for performing set operations on two given vectors. 
This implies that we can apply functions such as set union, intersection, 
difference, equality and membership, on `"character"` vectors. 

| Function       | Description    |
|:---------------|:---------------|
| `union()`      | set union      |
| `intersect()`  | intersection   |
| `setdiff()`    | set difference |
| `setequal()`   | equal sets     |
| `identical()`  | exact equality |
| `is.element()` | is element     |
| `%in%()`     | contains       |
| `sort()`       | sorting        |


### Set union with `union()`

Let's start our reviewing of set functions with `union()`. As its name 
indicates, you can use `union()} when you want to obtain the elements of 
the union between two character vectors:

```{r union_ex}
# two character vectors
set1 <- c("some", "random", "words", "some")
set2 <- c("some", "many", "none", "few")

# union of set1 and set2
union(set1, set2)
```

Notice that `union()` discards any duplicated values in the provided vectors. 
In the previous example the word `"some"` appears twice inside `set1` but it 
appears only once in the union. In fact all the set operation functions will 
discard any duplicated values.


### Set intersection with `intersect()`

Set intersection is performed with the function `intersect()`. You can use this
function when you wish to get those elements that are common to both vectors:

```{r intersect_ex}
# two character vectors
set3 <- c("some", "random", "few", "words")
set4 <- c("some", "many", "none", "few")

# intersect of set3 and set4
intersect(set3, set4)
```


### Set difference with `setdiff()`

Related to the intersection, you might be interested in getting the difference 
of the elements between two character vectors. This can be done with `setdiff()`:

```{r setdiff_ex}
# two character vectors
set5 <- c("some", "random", "few", "words")
set6 <- c("some", "many", "none", "few")

# difference between set5 and set6
setdiff(set5, set6)
```


### Set equality with `setequal()`

The function `setequal()` allows you to test the equality of two character 
vectors. If the vectors contain the same elements, `setequal()` returns `TRUE` (`FALSE` otherwise)

```{r setequal_ex}
# three character vectors
set7 <- c("some", "random", "strings")
set8 <- c("some", "many", "none", "few")
set9 <- c("strings", "random", "some")

# set7 == set8?
setequal(set7, set8)

# set7 == set9?
setequal(set7, set9)
```


### Exact equality with `identical()`

Sometimes `setequal()` is not always what we want to use. It might be the case 
that you want to test whether two vectors are _exactly equal_ (element by 
element). For instance, testing if `set7` is exactly equal to `set9`. Although 
both vectors contain the same set of elements, they are not exactly the same 
vector. Such test can be performed with the function `identical()`

```{r setidentical_ex}
# set7 identical to set7?
identical(set7, set7)

# set7 identical to set9?
identical(set7, set9)
```

If you consult the help documentation of `identical()`, you will see that this 
function is the "safe and reliable way to test two objects for being exactly 
equal".



### Element contained with `is.element()`

If you wish to test if an element is contained in a given set of character 
strings you can do so with `is.element()`:

```{r iselement_ex}
# three vectors
set10 <- c("some", "stuff", "to", "play", "with")
elem1 <- "play"
elem2 <- "crazy"

# elem1 in set10?
is.element(elem1, set10)

# elem2 in set10?
is.element(elem2, set10)
```

Alternatively, you can use the binary operator `%in%` to test if an element 
is contained in a given set. The function `%in%` returns `TRUE` if the first 
operand is contained in the second, and it returns `FALSE` otherwise:

```{r in_ex}
# elem1 in set10?
elem1 %in% set10

# elem2 in set10?
elem2 %in% set10
```


### Sorting with `sort()`

The function `sort()` allows you to sort the elements of a vector, either in 
increasing order (by default) or in decreasing order using the argument `decreasing`:

```{r sort_ex1}
set11 = c("today", "produced", "example", "beautiful", "a", "nicely")

# sort (decreasing order)
sort(set11)

# sort (increasing order)
sort(set11, decreasing = TRUE)
```

If you have alpha-numeric strings, `sort()` will put the numbers first when 
sorting in increasing order:

```{r sort_ex2}
set12 = c("today", "produced", "example", "beautiful", "1", "nicely")

# sort (decreasing order)
sort(set12)

# sort (increasing order)
sort(set12, decreasing = TRUE)
```
