# (PART) Characters {-}

# Character Strings in R {#chars}

## Introduction

This chapter introduces you to the basic concepts for creating character
vectors and character strings in R. You will also learn how R treats objects 
containing characters.

In R, a piece of text is represented as a sequence of characters (letters, 
numbers, and symbols). The data type R provides for storing sequences of 
characters is _character_. Formally, the __mode__ of an object that holds 
character strings in R is `"character"`.

You express character strings by surrounding text within double quotes:

```{r eval=FALSE}
"a character string using double quotes"
```

or you can also surround text within single quotes:

```{r eval=FALSE}
'a character string using single quotes'
```

The important thing is that you must match the type of quotes that your are 
using. A starting double quote must have an ending double quote. Likewise, a 
string with an opening single quote must be closed with a single quote. 

Typing characters in R like in above examples is not very useful. Typically, 
you are going to create objects or variables containing some strings. For 
example, you can create a variable `string` that stores some string:

```{r}
string <- 'do more with less'
string
```

Notice that when you print a character object, R displays it using double 
quotes (regardless of whether the string was created using single or double 
quotes). This allows you to quickly identify when an object contains character 
values.

When writing strings, you can insert single quotes in a string with double 
quotes, and vice versa:

```{r}
# single quotes within double quotes
ex1 <- "The 'R' project for statistical computing"
```

```{r}
# double quotes within single quotes
ex2 <- 'The "R" project for statistical computing'
```

However, you cannot directly insert single quotes in a string with single 
quotes, neither you can insert double quotes in a string with double quotes 
(Don't do this!):

```{r eval=FALSE}
ex3 <- "This "is" totally unacceptable"
```

```{r eval=FALSE}
ex4 <- 'This 'is' absolutely wrong'
```

In both cases R will give you an error due to the unexpected presence of 
either a double quote within double quotes, or a single quote within single 
quotes.

If you really want to include a double quote as part of the string, you need 
to _escape_ the double quote using a backslash `\` before it:

```{r eval = FALSE}
"The \"R\" project for statistical computing"
```

We will talk more about escaping characters in the following chapters.


## Common use of strings in R

Perhaps the most common use of character strings in R has to do with:

- names of files and directories
- names of elements in data objects
- text elements displayed in plots and graphs
  
When you read a file, for instance a data table stored in a csv file, 
you typically use the `read.table()` function and friends---e.g. `read.csv()`, 
`read.delim()`. Assuming that the file `dataset.csv` is in your working directory:

```{r eval = FALSE}
dat <- read.csv(file = 'dataset.csv')
```

The main parameter for the function `read.csv()` is `file` which requires a 
character string with the pathname of the file.

Another example of a basic use of characters is when you assign names to the 
elements of some data structure in R. For instance, if you want to name the 
elements of a (numeric) vector, you can use the function `names()` as follows:

```{r eval = FALSE}
num_vec <- 1:5
names(num_vec) <- c('uno', 'dos', 'tres', 'cuatro', 'cinco')
num_vec
```

Likewise, many of the parameters in the plotting functions require some sort 
of input string. Below is a hypothetical example of a scatterplot that includes 
several graphical elements like the main title (`main`), subtitle (`sub`), 
labels for both x-axis and y-axis (`xlab`, `ylab`), the name of the color
(`col`), and the symbol for the point character (`pch`).

```{r eval = FALSE}
plot(x, y, 
     main = 'Main Title', 
     sub = 'Subtitle',
     xlab = 'x-axis label', 
     ylab = 'y-axis label',
     col = 'red', 
     pch = 'x')
```


## Creating Character Strings

Besides the single quotes `''` or double quotes `""`, R provides the function 
`character()` to create character vectors. More specifically, `character()` 
is the function that creates vector objects of type `"character"`. 

When using `character()` you just have to specify the length of the vector. 
The output will be a character vector filled of empty strings:

```{r character_function}
# character vector with 5 empty strings
char_vector <- character(5)
char_vector
```

When would you use `character()`? A typical usage case is when you want to 
initialize an _empty_ character vector of a given length. The idea is to 
create an object that you will modify later with some computation.

As with any other vector, once an empty character vector has been created, 
you can add new components to it by simply giving it an index value outside 
its previous range: 

```{r empty_vector_ex1}
# another example
example <- character(0)
example

# check its length
length(example)

# add first element
example[1] <- "first"
example

# check its length again
length(example)
```

You can add more elements without the need to follow a consecutive index range:

```{r empty_vector_ex2}
example[4] <- "fourth"
example
length(example)
```

Notice that the vector `example` went from containing one-element to contain 
four-elements without specifying the second and third elements. R fills this 
gap with missing values `NA`.


### Empty string

The most basic type of string is the __empty string__ produced by 
consecutive quotation marks: `""`. Technically, `""` is a string with 
no characters in it, hence the name "empty string":

```{r empty_string}
# empty string
empty_str <- ""
empty_str

# class
class(empty_str)
```


### Empty character vector

Another basic string structure is the __empty character vector__ produced 
by the function `character()` and its argument `length=0`:

```{r empty_char_vector}
# empty character vector
empty_chr <- character(0)
empty_chr

# class
class(empty_chr)
```

It is important not to confuse the empty character vector `character(0)` 
with the empty string `""`; one of the main differences between them is 
that they have different lengths:

```{r empty_str_char_lengths}
# length of empty string
length(empty_str)

# length of empty character vector
length(empty_chr)
```

Notice that the empty string `empty_str` has length 1, while the empty 
character vector `empty_chr` has length 0.

Also, `character(0)` occurs when you have a character vector with one or 
more elements, and you attempt to subset the position 0:

```{r sunny}
string <- c('sun', 'sky', 'clouds')
string
```

If you try to retrieve the element in position 0 you get:

```{r sunny0}
string[0]
```


### Function `c()`

There is also the generic function `c()` (concatenate or combine) that you 
can use to create character vectors. Simply pass any number of character 
elements separated by commas:

```{r}
string <- c('sun', 'sky', 'clouds')
string
```

Again, notice that you can use single or double quotes to define the character 
elements inside `c()`

```{r}
planets <- c("mercury", 'venus', "mars")
planets
```



### `is.character()` and `as.character()`

Related to `character()` R provides two related functions: 
`as.character()` and `is.character()`. These two functions are methods for 
coercing objects to type `"character"`, and testing whether an 
R object is of type `"character"`. For instance, let's define two 
objects `a` and `b` as follows:

```{r objs_a_b}
# define two objects 'a' and 'b'
a <- "test me"
b <- 8 + 9
```

To test if `a` and `b` are of type `"character"` use the function
`is.character()`:

```{r is_character}
# are 'a' and 'b' characters?
is.character(a)

is.character(b)
```

Likewise, you can also use the function `class()` to get the class of an 
object:

```{r class_a_b}
# classes of 'a' and 'b'
class(a)
class(b)
```

The function `as.character()` is a coercing method. For better or worse, 
R allows you to convert (i.e. coerce) non-character objects into character 
strings with the function `as.character()`:

```{r as_character}
# converting 'b' as character
b <- as.character(b)
b
```


## Behavior of R objects with character strings

The main, and most basic, type of objects in R are vectors. Vectors must have 
their values _all of the same mode_. This means that any given vector 
must be unambiguously either logical, numeric, complex, character or raw. 
In R we say that vectors are __atomic__ structures, with their elements 
having all the same type or mode. 

So what happens when you mix different types of data in a vector?

```{r mixed_vector}
# vector with numbers and characters
c(1:5, pi, "text")
```

As you can tell, the resulting vector from combining integers `1:5`, the 
number `pi`, and some `"text"` is a vector with all its elements 
treated as character strings. In other words, when you combine mixed data in 
vectors, strings will dominate. This means that the mode of the vector will be 
`"character"`, even if you mix logical values:

```{r mixed_vector2}
# vector with numbers, logicals, and characters
c(1:5, TRUE, pi, "text", FALSE)
```

In fact, R follows two basic rules of data types coercion. The most strict 
rule is: if a character string is present in a vector, everything else in the 
vector will be converted to character strings. The other coercing rule is: if a 
vector only has logicals and numbers, then logicals will be converted to 
numbers; `TRUE` values become 1, and `FALSE` values become 0.

Keeping these rules in mind will save you from many headaches and frustrating 
moments. Moreover, you can use them in your favor to manipulate data in very 
useful ways.


__Matrices.__
The same behavior of vectors happens when you mix characters and numbers in 
matrices. Again, everything will be treated as characters:

```{r mixed_matrix}
# matrix with numbers and characters
rbind(1:5, letters[1:5])
```

__Data frames.__
With data frames, things are a bit different. By default, character strings 
inside a data frame will be converted to factors:

```{r mixed_dataframe1}
# data frame with numbers and characters
df1 = data.frame(numbers=1:5, letters=letters[1:5])
df1
# examine the data frame structure
str(df1)
```

To turn-off the `data.frame()`'s default behavior of converting strings 
into factors, use the argument `stringsAsFactors = FALSE`:

```{r mixed_dataframe2, tidy=FALSE}
# data frame with numbers and characters
df2 <- data.frame(
  numbers = 1:5, 
  letters = letters[1:5], 
  stringsAsFactors = FALSE)

df2
# examine the data frame structure
str(df2)
```

Even though `df1` and `df2` are identically displayed, their structure is 
different. While `df1$letters` is stored as a `"factor"`, `df2$letters` is 
stored as a `"character"`.

__Lists.__
With lists, you can combine any type of data objects. The type of data in 
each element of the list will maintain its corresponding mode:

```{r mixed_list}
# list with elements of different mode
list(1:5, letters[1:5], rnorm(5))
```




## The workhorse function `paste()`

The function `paste()` is perhaps one of the most important functions that you
can use to create and build strings. `paste()` takes one or more R objects, 
converts them to `"character"`, and then it concatenates (pastes) them to form 
one or several character strings. Its usage has the following form:

```
paste(..., sep = " ", collapse = NULL)
```

The argument `...` means that it takes any number of objects. The argument 
`sep` is a character string that is used as a separator. The argument 
`collapse` is an optional string to indicate if you want all the terms to be 
collapsed into a single string. Here is a simple example with `paste()`:

```{r paste_string_ex1}
# paste
PI <- paste("The life of", pi)

PI
```

As you can see, the default separator is a blank space (`sep = " "`). But you 
can select another character, for example `sep = "-"`:

```{r paste_string_ex2}
# paste
IloveR <- paste("I", "love", "R", sep = "-")

IloveR
```

If you give `paste()` objects of different length, then it will apply a 
recycling rule. For example, if you paste a single character `"X"` with the
sequence `1:5`, and separator `sep = "."`, this is what you get:

```{r paste_string_ex3}
# paste with objects of different lengths
paste("X", 1:5, sep = ".")
```

To see the effect of the `collapse` argument, let's compare the difference with
collapsing and without it:

```{r paste_string_ex4}
# paste with collapsing
paste(1:3, c("!","?","+"), sep = '', collapse = "")

# paste without collapsing
paste(1:3, c("!","?","+"), sep = '')
```

One of the potential problems with `paste()` is that it coerces missing values `NA` into the character `"NA"`:

```{r paste_string_ex5}
# with missing values NA
evalue <- paste("the value of 'e' is", exp(1), NA)

evalue
```

In addition to `paste()`, there's also the function `paste0()` which is the equivalent of

```
paste(..., sep = "", collapse)
```

```{r paste0}
# collapsing with paste0
paste0("let's", "collapse", "all", "these", "words")
```


## Exercises

1. What is the difference between the empty character `""` and the output 
of invoking `character()`?

2. When you combine logical values, numeric values, and character
values in one single vector, what will be the mode of the resulting vector?

3. Using `rep()`, how would you obtain the following character vectors:

```{r echo = FALSE}
rep(c("a", "b"), times = 4)
rep(c("a", "b"), each = 4)
rep(c("a", "b"), each = 2, times = 2)
```

4. Given the following vectors `go`, `bears`, and `bang`:

```{r}
go <- 'Go'
bears <- 'Bears'
bang <- '!'
```

how would you use `paste()` and `paste0()` to get the following strings:

```
"Go Bears !"
"GoBears!"
"Go Bears!"
"Go-Bears!"
"Go Bears!!!"
"Go Bears! Go Bears! Go Bears!"
```
