# Input and Output



## Output

Some times you need to export results to a file. Typically this happens when
you want to export a data table to a text file. R provides the functions
`write.table()` and `write.csv()` for these purposes.
These functions let you send a matrix or a data frame to a text file that will
have a tabular format (i.e. rows and columns).


### Concatenating output

You can use `cat()` to concatenate and print information to a file. 

To show you how to use `cat()` let's illustrate a simple example using the data 
frame `mtcars` that comes in R.

```{r eval = FALSE}
# summary statistics of unemp
min(mtcars$mpg)
max(mtcars$mpg)
median(mtcars$mpg)
mean(mtcars$mpg)
sd(mtcars$mpg)
```

The goal is to generate a file `mpg-statistics.txt` with the following 
contents:

```
Miles per Gallon Statistics

Minimum: 10.40
Maximum: 33.90
Median : 19.20
Mean   : 20.09 
Std Dev: 6.02
```

Here is one way to do it. First, let's assign the statistics to different
objects:

```{r}
# summary statistics of mpg
mpg_min <- min(mtcars$mpg)
mpg_max <- max(mtcars$mpg)
mpg_med <- median(mtcars$mpg)
mpg_avg <- mean(mtcars$mpg)
mpg_sd <- sd(mtcars$mpg)
```

After creating the objects containing the summary statistics, the next step
is to export them to the text file `mpg-statistics.txt` via `cat()`. 
Assuming that the output file is in your working directory, here's how you can
send the set of strings to the text file:

```{r eval = FALSE}
# name of output file
outfile <- "mpg-statistics.txt"

# first line of the file
cat("Miles per Gallon Statistics\n\n", file = outfile)

# subsequent lines appended to the output file
cat("Minimum:", mpg_min, "\n", file = outfile, append = TRUE)
cat("Maximum:", mpg_max, "\n", file = outfile, append = TRUE)
cat("Median :", mpg_med, "\n", file = outfile, append = TRUE)
cat("Mean   :", mpg_avg, "\n", file = outfile, append = TRUE)
cat("Std Dev:", mpg_sd, "\n", file = outfile, append = TRUE)
```

The first line exported to `mpg-statistics.txt` is a string with the title
`"Miles per Gallon Statistics\n\n"`. Observe that we are using two new line
characters `"\n\n"` to add some space between the title and the statistics.
The rest of calls to `cat()` use the argument `append = TRUE` to concatenate
the specified strings to the end of the text file without overriding the 
existing lines.

If you run the code of this example and look at the contents of 
`mpg-statistics.txt`, you will see the following output:

```
Miles per Gallon Statistics

Minimum: 10.4 
Maximum: 33.9 
Median : 19.2 
Mean   : 20.09062 
Std Dev: 6.026948 
```

As you can tell, the displayed values have a different number of decimal 
digits. If you just want to keep two decimal digits, you can use `sprintf()`
and choose the format `"%0.2f"`. Let's re-export the lines:

```{r eval=FALSE}
cat("Miles per Gallon Statistics\n\n", file = outfile)
cat(sprintf('Minimum: %0.2f', mpg_min), "\n", file = outfile, append = TRUE)
cat(sprintf('Maximum: %0.2f', mpg_max), "\n", file = outfile, append = TRUE)
cat(sprintf('Median : %0.2f', mpg_med), "\n", file = outfile, append = TRUE)
cat(sprintf('Mean   : %0.2f', mpg_avg), "\n", file = outfile, append = TRUE)
cat(sprintf('Std Dev: %0.2f', mpg_sd), "\n", file = outfile, append = TRUE)
```

Now the content of `mpg-statistics.txt` should look like this:

```
Miles per Gallon Statistics

Minimum: 10.40 
Maximum: 33.90 
Median : 19.20 
Mean   : 20.09 
Std Dev: 6.03 
```

Here is an exercise for you: How would you avoid writing that many calls to 
`cat()`?


### Sinking output

Another interesting function is `sink()`. This function is very useful when
you want to export R output as is displayed in the R console. For example, 
consider the output from `summary()`

```{r eval=FALSE}
summary(mtcars)
```

You could assign the output of `summary(mtcars)` to an object and then try
`writeLines()` to export the results to a file `mtcars-summary.txt`, but 
you won't keep the same format of R:

```{r eval = FALSE}
mtcars_summary <- summary(mtcars)
writeLines(mtcars_summary, con = "mtcars-summary.txt")
```

To be able to keep the same output display of R, you must use `sink()`. This
function will __divert__ R output to the specified file:

```{r eval=FALSE}
sink(file = "mtcars-statistics.txt")
summary(australia)
sink()
```

__Your turn:__ Use `sink()` to send the output from running a linear 
regression of `mpg` on `disp` with the function `lm()`. Also export the results 
from using `summary()` on the regression object. And/or try running a t-test
between `mpg` and `disp` with `t.test()`.



## Exporting tables

Another interesting tool to export tables in LaTeX or HTML formats is provided
by the R package `"xtable"` and its main function `xtable()`.

```{r eval = FALSE}
library(xtable)

# linear regression
reg <- lm(mpg ~ disp, data = mtcars)

# create xtable and export it
reg_table <- xtable(reg)
```

The object `reg_table` is an object of class `"xtable"`. What you do with this
type of objects is `print()` them to a file.

To print `reg_table` in latex format to a `.tex` file:

```{r eval = FALSE}
print(reg_table, type = "latex", file = "reg-table.tex")
```

To print `reg_table` in html format to an `.html` file:

```{r eval = FALSE}
print(reg_table, type = "html", file = "reg-table.html")
```
