# Data Log File {#logs}

In this example, we'll be using the text file `logfile.txt` located in the 
`data/` folder of the book's github repository:

[https://raw.githubusercontent.com/gastonstat/r4strings/master/data/logfile.txt](https://raw.githubusercontent.com/gastonstat/r4strings/master/data/logfile.txt)

This file is a __server log file__ that contains the recorded events taking place in a web server. The content of the file is in a special format known as _common log format_. According to [wikipedia](https://en.wikipedia.org/wiki/Common_Log_Format): 

> "The Common Log Format is a standardized text file format used by web servers when generating server log files."

Here's an example of a log record; the text should in one line of code, but I've split it into 2 lines for readibility purposes:

```
pd9049dac.dip.t-dialin.net - - [01/May/2001:01:51:25 -0700] 
"GET /accesswatch/accesswatch-1.33/ HTTP/1.0" 200 1004
```

- A `"-"` in a field indicates missing data.
- `pd9049dac.dip.t-dialin.net` is the IP address of the client (remote host) which made the request to the server.
- `[01/May/2001:01:51:25 -0700]` is the date, time, and time zone that the request was received, by default in strftime format` %d/%b/%Y:%H:%M:%S %z`.
- `"GET /accesswatch/accesswatch-1.33/ HTTP/1.0"` is the request line from the client. 
- The method `GET, /accesswatch/accesswatch-1.33/` is the resource requested, and `HTTP/1.0` is the HTTP protocol.
- `200` is the HTTP status code returned to the client. 
    + `2xx` is a successful response
    + `3xx` a redirection
    + `4xx` a client error, and 
    + `5xx` a server error
- `1004` is the size of the object returned to the client, measured in bytes.

If you want to download a copy of the text file to your working directory run the following code:

```{r eval = FALSE}
# download file
github <- "https://raw.githubusercontent.com/gastonstat/r4strings"
textfile <- "/master/data/logfile.txt"
download.file(url = paste0(github, textfile), destfile = "logfile.txt")
```


```{r may-logs, echo = FALSE}
# one option is to read in the content with 'readLines()'
logs <- readLines('data/logfile.txt')
```


## Reading the text file

The first step involves reading the data in R. How can you do this? One option is with the `readLines()` function which reads any text file into a character vector:

```{r may-logs-fake, eval = FALSE}
# one option is to read in the content with 'readLines()'
logs <- readLines('logfile.txt')
```

Let's take a peek at the content of the vector `logs`:

```{r head-logs}
# take a peek at the contents in logs
head(logs)
```

Because the file contains `r length(logs)` lines (or elements), let's get a subset by taking a random sample of size 50:

```{r}
# subset a sample of lines
set.seed(98765)
s <- sample(1:length(logs), size = 50)
sublogs <- logs[s]
```


### JPG File Requests

To begin our regex experiments, let's try to find out "how many requests involved a JPG file?".
One way to answer the previous question is by counting the number of lines containing the pattern `"jpg"`. We can use `grep()` to match or detect this pattern:

```{r}
# matching "jpg" (which lements)
grep("jpg", sublogs)

# showing value of matches
grep("jpg", sublogs, value = TRUE)
```

We can try to be more specific by defining a pattern `".jpg"` in which the `.` corresponds to the _literal_ dot character. To match the dot, we need to escape it with `"\\."`:

```{r}
# we could try to be more precise and match ".jpg"
grep("\\.jpg ", sublogs, value = TRUE)
```

A similar output of `grep()` can be obtained with `str_detect()`, which allows you to _detect_ what elements contain a match to the specified pattern:

```{r}
# matching "jpg" (which lements)
str_detect(string = sublogs, pattern = "\\.jpg")
```

We can do the same for PNG extensions (or for GIF or ICO):

```{r}
# matching "png" (which lements)
str_detect(string = sublogs, pattern = "\\.png")
```


### Extracting file extensions

Another common task when working with regular expressions has to do with pattern extraction. For this purposes, we can use `str_extract()`:

```{r}
# extracting "jpg" (which lements)
str_extract(string = sublogs, pattern = "\\.jpg")
```

`str_extract()` actually let us confirm that we are matching the desired patterns. Notice that when there is no match, `str_extract()` returns a missing value `NA`.


### Image files

Now let's try to detect all types of image files: JPG, PNG, GIF, ICO

```{r}
# looking for image file extensions
jpgs <- str_detect(logs, pattern = "\\.jpg ")
sum(jpgs)

pngs <- str_detect(logs, pattern = "\\.png ")
sum(pngs)

gifs <- str_detect(logs, pattern = "\\.gif")
sum(gifs)

icos <- str_detect(logs, pattern = "\\.ico ")
sum(icos)
```


### How to match image files with one regex pattern?

We can use character sets to define a more generic pattern. For instance, to
match `"jpg"` or `"png"`, we could join three character sets: `"[jp][pn][g]"`.
The first set `[jp]` looks for `j` or `p`, the second set `[pn]` looks for
`p` or `n`, and the third set simply looks for `g`.

```{r}
# matching "jpg" or "png"
jpg_png_lines <- str_detect(sublogs, "[jp][pn][g]")
sum(jpg_png_lines)
```

Including the dot, we can use: `"\\.[jp][pn][g]"`

```{r}
# matching "jpg" or "png"
jpg_png_lines <- str_detect(sublogs, "\\.[jp][pn][g]")
sum(jpg_png_lines)
```

We could generalize the pattern to include the GIF and ICO extensions:

```{r}
# matching "jpg" or "png" or "gif"
image_lines1 <- str_detect(sublogs, "[jpgi][pnic][gfo]")
sum(image_lines1)
```

To confirm that we are actually matching `jpg`, `png`, `gif` and `ico`, let's
use `str_extract()`

```{r}
# are we correctly extracting image file extensions?
str_extract(sublogs, "[jpgi][pnic][gfo]")
```

The previous pattern does not really work as expected: note that we are matching 
the patterns formed by `"ing"` and `"inf"` which do not correspond to image file 
extensions.

An alternative way to detect JPG and PNG is by grouping patterns inside
parentheses, and separating them with the metacharacter `"|"` which means _OR_:

```{r}
# detecting .jpg OR .png
jpg_png <- str_detect(sublogs, "\\.jpg|\\.png")
sum(jpg_png)
```

Here's how to detect all the extension in one single pattern:

```{r}
# matching "jpg" or "png" or "gif" or "ico"
image_lines <- str_detect(sublogs, "\\.jpg|\\.png|\\.gif|\\.ico")
sum(image_lines)
```

To make sure our regex operation is successful, let's see the output of 
`str_extract()`:

```{r}
images_output <- str_extract(sublogs, "\\.jpg|\\.png|\\.gif|\\.ico")
images_output
```

There's some repetition with the dot character; we can modify our previous 
pattern by placing the dot `"\\."` at the beginning:

```{r}
images_output <- str_extract(sublogs, "\\.jpg|png|gif|ico")
images_output
```

Notice that the dot only appears next to `".jpg"` but not with the other 
type of extensions. What we need to do is group the file extensions by surrounding
them with parentheses:

```{r}
images_output <- str_extract(sublogs, "\\.(jpg|png|gif|ico)")
images_output
```

Now let's apply the pattern on the entire log file, to count the number of files 
of each type:

```{r}
# frequencies
img_extensions <- str_extract(logs, "\\.(jpg|png|gif|ico)")
table(img_extensions)
```
