# (PART) Formatting {-}

# Formatting Text and Numbers {#formatting}

## Introduction

A common task when working with character strings involves printing and 
displaying them on the screen or on a file. In this chapter you will learn 
about the different functions and options in R to print strings in a 
wide variety of common---and not so common---formats. 


## Printing and formatting

R provides a series of functions for printing strings. Some of the printing 
functions are useful when creating `print` methods for programmed 
objects' classes. Other functions are useful for printing output either in the 
R console or to a given file. In this chapter we will describe the following 
print-related functions:

| Function     | Description          |
|:-------------|:---------------------|
| `print()`    | generic printing     |
| `noquote()`  | print with no quotes |
| `cat()`      | concatenation        |
| `format()`   | special formats      |
| `toString()` | convert to string    |
| `sprintf()`  | C-style printing     |



## Generic printing

The _workhorse_ printing function in R is `print()`. As its 
names indicates, this function prints its argument on the R console:

```{r print_string}
# text string
my_string <- "programming with data is fun"

print(my_string)
```

To be more precise, `print()` is a generic function, which means that you
should use this function when creating printing methods for programmed classes.

As you can see from the previous example, `print()` displays text in 
quoted form by default. If you want to print character strings with no quotes 
you can set the argument `quote = FALSE` 

```{r print_noquote}
# print without quotes
print(my_string, quote = FALSE)
```


### When to use `print()`?

When you type the name of an obbject in the R console, R calls the corresponding
`print` method associated to the class of the object. If the object is a 
`"data.frame"`, then R will dispatch the method `print.data.frame` and display 
the output on screen accordingly.

Most of the times you don't really need to invoke `print()`. Usually, simply 
typing the name of the object will suffice. So when do you actually call 
`print()`? You use `print()` when your code is inside an __R expression__ 
(i.e. code inside curly braces `{ }`) and you want to see the results of one
or more computational steps. Typical examples that require an explicit call to
`print()` is when you are interested in looking at some value within a loop, 
or a conditional structure.

Consider the following dummy `for` loop. It iterates five times, each 
time adding 1 to the value of the iterator `i`:

```{r}
for (i in 1:5) {
  i + 1
}
```

The above code works and R executes the additions, but nothing is displayed on 
the console. This is because the command `i + 1` forms part of an R expression,
that is, it is within the braces `{ }`.
To be able to see the actual computations you should call `print()` like so:

```{r}
for (i in 1:5) {
  print(i + 1)
}
```


## Concatenate and print with `cat()`

Another very useful function is `cat()` which allows you to concatenate objects 
and print them either on screen or to a file. Its usage has the following 
structure:

```
cat(..., file = "", sep = " ", fill = FALSE, labels = NULL, append = FALSE)
```

The argument `...` implies that `cat()` accepts several types of R objects 
(typically vectors). However, when we pass numeric and/or complex vectors, 
they are automatically converted to character strings by `cat()`. By default, 
the strings are concatenated with a space character as separator. This can be 
modified with the `sep` argument.

If you use `cat()` with only one single string, you get a similar 
(although not identical) result as `noquote()`:

```{r cat_ex1}
# simply print with 'cat()'
cat(my_string)
```

As you can see, `cat()` prints its arguments without quotes. In essence, `cat()` 
simply displays its content (on screen or in a file). Compared to `noquote()`, 
`cat()` does not print the numeric line indicator (`[1]` in this case). 

The usefulness of `cat()` is when we have two or more strings that we want 
to concatenate:

```{r cat_ex2}
# concatenate and print
cat(my_string, "with R")
```

You can use the argument `sep` to indicate a chacracter vector that will be 
included to separate the concatenated elements:

```{r cat_ex3}
# especifying 'sep'
cat(my_string, "with R", sep=" =) ")

# another example
cat(1:10, sep = "-")
```

When we pass vectors to `cat()`, each of the elements are treated as though 
they were separate arguments:

```{r cat_ex4}
# first four months
cat(month.name[1:4], sep = " ")

# first four months
cat(month.name[1:4], sep = "-")

# first four months
cat(month.name[1:4], sep = "")
```

The argument `fill` allows us to break long strings; this is achieved when we 
specify the string width with an integer number:

```{r cat_ex5}
# fill = 30
cat("Loooooooooong strings", "can be displayed", "in a nice format", 
    "by using the 'fill' argument", fill = 30)
```

Last but not least, we can specify a file output in `cat()`. For instance, 
let's suppose that we want to save the output in the file `output.txt` 
located in our working directory. This is done by specifying the `file`
argument:

```{r cat_ex6, eval=FALSE}
# cat with output in a given file
cat(my_string, "with R", file = "output.txt")
```




## Encoding strings with `format()`

The function `format()` allows you to format an R object for _pretty_ printing. 
Essentially, `format()` treats the elements of a vector as character strings 
using a common format. This is especially useful when printing numbers and 
quantities under different formats.

```{r format_string_default}
# default usage
format(13.7)

# another example
format(13.12345678)
```


Some useful arguments used in `format()`:

- `width` the (minimum) width of strings produced
- `trim` if set to `TRUE` there is no padding with spaces
- `justify` controls how padding takes place for strings. 
Takes the values `"left", "right", "centre", "none"`


For controling the printing of numbers, use these arguments:

- `digits` The number of digits to the right of the decimal place.
- `scientific` use `TRUE` for scientific notation, `FALSE` for standard notation


```{r format_string}
# use of 'nsmall'
format(13.7, nsmall = 3)

# use of 'digits'
format(c(6.0, 13.1), digits = 2)

# use of 'digits' and 'nsmall'
format(c(6.0, 13.1), digits = 2, nsmall = 1)
```

By default, `format()` pads the strings with spaces so that they all have 
the same length.

```{r format_strings_ihaka}
# justify options
format(c("A", "BB", "CCC"), width = 5, justify = "centre")

format(c("A", "BB", "CCC"), width = 5, justify = "left")

format(c("A", "BB", "CCC"), width = 5, justify = "right")

format(c("A", "BB", "CCC"), width = 5, justify = "none")

# digits
format(1/1:5, digits = 2)

# use of 'digits', widths and justify
format(format(1/1:5, digits = 2), width = 6, justify = "c")
```

For printing large quantities with a sequenced format we can use the arguments 
`big.mark` or `big.interval`. For instance, here is how we can print a number 
with sequences separated by a comma `","`

```{r format_bigmark}
# big.mark
format(123456789, big.mark = ",")
```


## Exercises

1. Why do we say that `print()` is not really one function but a _family_ of functions?

2. Mention three differences between `print()` and `cat()`.

3. What happens when you pass a matrix to `cat()`? For instance:

```{r cat_matrix, eval = FALSE}
m <- matrix(1:12, nrow = 3, ncol = 4)
cat(m)
```

4. What happens when you pass a data frame object to `cat()`? For instance:

```{r cat_dataframe, eval = FALSE}
dfr <- data.frame(a = 1, b = 2, c = 3)
cat(dfr)
```
