# (PART) Regex {-}

# Regular Expressions {#regex1}

## Introduction

So far you have learned some basic and intermediate functions for handling and 
working with text in R. These are very useful functions and they allow you 
to do many interesting things. However, if you truly want to unleash the power 
of strings manipulation, you need to take things to the next level and learn 
about _regular expressions_.


## What are Regular Expressions?

The name "Regular Expression" does not say much. However, regular 
expressions are all about text. Think about how much text is all around 
you in our modern digital world: email, text messages, news articles, blogs, 
computer code, contacts in your address book---all these things are text. 
Regular expressions are a tool that allows us to work with these text by 
describing text patterns.

A __regular expression__ is a special text string for describing a certain 
amount of text. This "certain amount of text" receives the formal name of 
_pattern_. In other words, a regular expression is a set of symbols that 
describes a text pattern. More formally we say that a regular expression is a 
_pattern that describes a set of strings_. In addition to this first 
meaning, the term regular expression can also be used in a slightly different 
but related way: as the formal language of these symbols that needs 
to be interpreted by a regular expression processor.
Because the term "regular expression" is rather long, most people 
use the word __regex__ as a shortcut term. And you will even find the 
plural _regexes_.

It is also worth noting what regular expressions are not. They're not a 
programming language. They may look like some sort of programming language
because they are a formal language with a defined set of rules that gets a 
computer to do what we want it to do. However, there are no variables in regex
and you can't do computations like adding 1 + 1. 


### What are Regular Expressions used for?

We use regular expressions to work with text. Some of its common uses involve
testing if a phone number has the correct number of digits, if a date follows a 
specifc format (e.g. mm/dd/yy), if an email address is in a valid format, or if 
a password has numbers and special characters. You could also use regular 
expressions to search a document for _gray_ spelt either as "gray" or 
"grey". You could search a document and replace all occurrences of "Will", "Bill", 
or "W." with William. Or you could count the number of times in a document 
that the word "analysis" is immediately preceded by the words "data", 
"computer" or "statistical" only in those cases. You could use it to 
convert a comma-delimited file into a tab-delimited file or to find duplicate 
words in a text. 

In each of these cases, you are going to use a regular expression to write up a 
description of what you are looking for using symbols. In the case of a phone 
number, that pattern might be three digits followed by a dash, followed by three 
digits and another dash, followed by four digits. Once you have defined a 
pattern then the regex processor will use our description to return matching 
results, or in the case of the test, to return true or false for whether or not 
it matched.


### A word of caution about regex

If you have never used regular expressions before, their syntax may seem a bit
scary and cryptic. You will see strings formed by a bunch of letters, digits, 
and other punctuation symbols combined in seemingly nonsensical ways. As with
any other topic that has to do with programming and data analysis, learning
the principles of regex and becoming fluent in defining regex patterns takes
time and requires a lot of practice. The more you use them, the better you
will become at defining more complex patterns and getting the most out of them.

Regular Expressions is a wide topic and there are books entirely dedicated to
this subject. The material offered in this book is not extensive and there are
many subtopics that I don't cover here. Despite the initial barriers that you
may encounter when entering the regex world, the pain and frustration of 
learning this tool will payoff in your data science career.


### Regular Expressions in R

Tools for working with regular expressions can be found in virtually all 
scripting languages (e.g. Perl, Python, Java, Ruby, etc). R has some 
functions for working with regular expressions but it does not provide 
the wide range of capabilities that other scripting languages do. Nevertheless, 
they can take you very far with some workarounds (and a bit of patience).

One of the best tools you must have in your toolkit is the R package `"stringr"`
(by Hadley Wickham). It provides functions that have similar behavior to 
those of the base distribution in R. But it also provides many more facilities
for working with regular expressions.


## Regex Basics

The main purpose of working with regular expressions is to describe patterns 
that are used to match against text strings. Simply put, working with regular 
expressions is nothing more than __pattern matching__. The result of a match 
is either successful or not.

The simplest version of pattern matching is to search for one occurrence (or 
all occurrences) of some specific characters in a string. For example, we might 
want to search for the word `"programming"` in a large text document, or we 
might want to search for all occurrences of the string `"apply"` in a series 
of files containing R scripts.

Typically, regular expression patterns consist of a combination of alphanumeric 
characters as well as special characters. A regex pattern can be as simple as a 
single character, or it can be formed by several characters with a more complex 
structure. In all cases we construct regular expressions much in the same form 
in which we construct arithmetic expressions, by using various operators to 
combine smaller expressions. 


## Matching Literal Characters

We're going to start with the simplest match of all: a literal character.
A literal character match is one in which a given character such as the letter
`"R"` matches the letter _R_. This type of match is the most basic 
type of regular expression operation: just matching plain text.

The following examples are extremely basic but they will help you get a 
good understanding of regex.

Consider the following text stored in a character vector `this_book`:

```{r book}
this_book <- 'This book is mine'
```

The first regular expression we are going to work with is `"book"`. 
This pattern is formed by a letter _b_, followed by a letter _o_, followed by 
another letter _o_, followed by a letter _k_. As you may guess, this pattern 
matches the word _book_ in the character vector `this_book`. 
To have a visual representation of the actual pattern that is matched, you
should use the function `str_view()` from the package `"stringr"` 
(you may need to upgrade to a recent version of RStudio):

```{r eval = FALSE}
str_view(this_book, 'book')
```

As you can tell, the pattern `"book"` doesn't match the entire content in
the vector `this_book`; it just matches those four letters.

It may seem really simple but there are a couple of details to be highlighted.
The first is that regex searches are case sensitive by default. This means 
that the pattern `"Book"` would not match _book_ in ``this_book``.

```{r eval = FALSE}
str_view("This Book is mine.", 'book')
```

You can change the matching task so that it is case insensitive but we will
talk about it later.

Let's add more text to `this_book`:

```{r book_online}
this_book <- 'This book is mine. I wrote this book with bookdown.'
```

Let's use `str_view()` to see what pieces of text are matched in `this_book` 
with the pattern `"book"`:

```{r grep_book, eval = FALSE}
str_view(this_book, "book")
```

As you can tell, only the first occurrence of _book_ was matched. This is
a common behavior of regular expressions in which they return a match as fast
possible. You can think of this behavior as the "eager principle", that is,
regular expressions are eager and they will give preference to an early match.
This is a minor but important detail and we will come back to this behavior of 
regular expressions.

All the letters and digits in the English alphabet are considered literal 
characters. They are called _literal_ because they match themselves.

```{r eval = FALSE}
str_view <- c("I had 3 quesadillas for lunch", "3")
```

Here is another example:

```{r}
transport <- c("car", "bike", "boat", "airplane")
```

The first pattern to test is the letter `"a"`:

```{r eval = FALSE}
str_view(transport, "a")
```

When you execute the previous command, you should be able to see that the 
letter `"a"` is highlighted in the words _car_, _boat_ and _airplane_.



## Metacharacters

The next topic that you should learn about regular expressions has to do with 
__metacharacters__. As you just learned, the most basic type of regular 
expressions are the literal characters which are characters that match 
themselves. However, not all characters match themselves. Any character that 
is not a literal character is a metacharacter. This type of characters have a 
special meaning and they allow you to transform literal characters in very 
powerful ways.

Below is the list of metacharacters in _Extended Regular Expressions_ (EREs):

```
.   \   |   (   )   [   ]   {   }   $   -    ^   *   +   ?
```

- the dot `.`
- the backslash `\`
- the bar `|`
- left or opening parenthesis `(`
- right or closing parenthesis `)`
- left or opening bracket `[`
- right or closing bracket `]`
- left or opening brace `{`
- right or closing brace `}`
- the dollar sign `$`
- the dash, hyphen or minus sign `-`
- the caret or hat `^`
- the star or asterisk `*`
- the plus sign `+`
- the question mark `?`

We're going to be working with these characters throughout the rest of the book. 
Simply put, everything else that you need to know about regular expressions 
besides literal characters is how these metacharacters work. The good news is 
that there are only a few metacharacters to learn. The bad news is 
that some metacharacters can have more than one meaning. And learning those 
meanings definitely takes time and requires hours of practice. The meaning of 
the metacharacters greatly depend on the context in which you use them,
how you use them, and where you use them. If it wasn't enough complication, 
it is also the metacharacters that have variation between the different 
regex engines. 
 


### The Wild Metacharacter

The first metacharacter you should learn about is the dot or period `"."`, 
better known as the __wild__ metacharacter. This metacharacter is used to
match __ANY__ character except for a new line.

For example, consider the pattern `"p.n"`, that is, _p wildcard n_. This
pattern will match _pan_, _pen_, and _pin_, but it will not match _prun_ 
or _plan_. The dot only matches one single character. 

Let's see another example using the vector `c("not", "note", "knot", "nut")` 
and the pattern `"n.t"`

```{r eval = FALSE}
not <- c("not", "note", "knot", "nut")

str_view(not, "n.t")
```

the pattern `"n.t"` matches _not_ in the first three elements, and _nut_
in the last element.

If you specify a pattern `"no."`, then just the first three elements
in `not` will be matched. 

```{r eval = FALSE}
str_view(not, "no.")
```

And if you define a pattern `"kn."`, then only the third element is matched.

```{r eval = FALSE}
str_view(not, "kn.")
```

The wild metacharacter is probably the most used metacharacter, and it is 
also the most abused one, being the source of many mistakes. Here is a basic
example with the regular expression formed by `"5.00"`. If you think that this 
pattern will match five with two decimal places after it, you will be 
surprised to find out that it not only matches _5.00_ but also _5100_ and _5-00_. 
Why? Because `"."` is the metacharacter that matches absolutely anything.
You will learn how to fix this mistake in the next section, but it illustrates 
an important fact about regular expressions: the challenge consists of matching 
what you want, but also in matching only what you want. You don't want to 
specify a pattern that is overly permissive. You want to find the thing you're 
looking for, but only that thing. 


### Escaping metacharacters

What if you just want to match the character dot? For example, say you 
have the following vector:

```{r}
fives <- c("5.00", "5100", "5-00", "5 00")
```

If you try the pattern `"5.00"`, it will match all of the elements in `fives`.

```{r eval = FALSE}
str_view(fives, "5.00")
```

To actually match the dot character, what you need to do is __escape__ the
metacharacter. In most languages, the way to escape a metacharacter is by 
adding a backslash character in front of the metacharacter: `"\."`. 
When you use a backslash in front of a metacharacter you are "escaping" the
character, this means that the character no longer has a special meaning, 
and it will match itself.

However, R is a bit different. Instead of using a backslash you have to use
two backslashes: `"5\\.00"`. This is because the backslash `"\"`, which is
another metacharacter, has a special meaning in R. Therefore, to match just
the element _5.00_ in `fives` in R, you do it like so:

```{r eval = FALSE}
str_view(fives, "5\\.00")
```
