# C-style Formatting

R comes with the `sprintf()` function that provides string formatting
like in the __C__ language. To be more precise, this function is a wrapper
for the C library function of the same name. In many other programming
languages, this type of printing is known as _printf_ which stands for 
__print formatting__. Simply put, 
`sprintf()` allows you to create strings as output using formatted data.

The function `sprintf()` requires using a special syntax that may look
awkward the first time you use it. Here is one example:

```{r}
sprintf("I woke up at %s:%s%s a.m.", 8, 0, 5)
```

How does `sprintf()` work? The first argument of this function is a character
vector of one element that contains the text to be formatted. Observe that 
inside the text there are various percent symbols `%` followed by the letter 
`s`. Each `%` is referred to as a __slot__, which is basically a placeholder 
for a variable that will be formatted. The rest of the inputs passed to
`sprintf()` are the values that will be used in each of the slots.

The string in the previous example contains three slots of the same type, `%s`,
and the subsequent arguments are numbers `8`, `0`, and `5`. Each number is used 
as a value for each slot. The letter `s` indicates that the formatted variable
is specified as a _string_.

Most of the times you won't use `sprintf()` like in the example above. 
Instead, what you will pass are variables containing different values:

```{r}
hour <- 8
mins1 <- 0
mins2 <- 5
sprintf("I woke up at %s:%s%s a.m.", hour, mins1, mins2)
```


## C-style formatting options

The string format `%s` is just one of a larger list of available formatting
options. The following table shows the most common formatting specifications:

| Notation | Description                                       |
|:---------|:--------------------------------------------------|
| `%s`     | a string                                          |
| `%d`     | an integer                                        |
| `%0xd`   | an integer padded with `x` leading zeros          |
| `%f`     | decimal notation with six decimals                |
| `%.xf`   | floating point number with `x` digits after decimal point | 
| `%e`     | compact scientific notation, `e` in the exponent  |
| `%E`     | compact scientific notation, `E` in the exponent  |
| `%g`     | compact decimal or scientific notation (with `e`) |



### Format Slot Syntax

The full syntax for a format slot is defined by:

```
%[parameter][flags][width][.precision][length]type
```

The percent symbol, `%`, as we said, indicates a placeholder or slot.


The `parameter` field is an optional field that can take the value `n$` in which 
`n` is the number of the variable to display, allowing the variables provided 
to be used multiple times, using varying format specifiers or in different 
orders.

```{r}
sprintf("The second number is %2$d, the first number is %1$d", 2, 1)
```


The `flags` field can be zero or more (in any order) of:

- `-` (minus) Left-align the output of this placeholder.
- `+` (plus) Prepends a plus for positive signed-numeric types.
- `' '` (space) Prepends a space for positive signed-numeric types.
- `0` (zero) When the 'width' option is specified, prepends zeros for numeric types.
- `#` (hash) Alternate form:
    + for `g` and `G` types, trailing zeros are not removed.
    + for `f, F, e, E, g, G` types, the output always contain a decimal point.
    + for `o, x, X` types, the text `0`, `0x`, `0X`, respectively, is 
    prepended to non-zero numbers.
    

The `width` field is an optional field that you use to specify a minimum number
of characters to output, and is typically used to pad fixed-width fields in 
tabulated output, where the fields would otherwise be smaller, although it 
does not cause truncation of oversized fields.

```{r}
sprintf("%*d", 5, 10)
```


The `precision` field usually specifies a maximum limit on the output, 
depending on the particular formatting type.

```{r}
sprintf("%.*s", 3, "abcdef")
```

The `length` field is also optional, and can be any of:


The most important field is the `type` field.

- `%`: Prints a literal `%` character (this type doesn't accept any flags, 
width, precision, length fields).
- `d, i`: integer value as signed decimal number.
- `f`: double value in normal fixed-point notation.
- `e, E`: double value in standard form.
- `g, G`: double value in either normal or exponential notation.
- `x, X`: unsigned integer as a hexadecimal number. `x` uses lower case, while
`X` uses upper case.
- `o`: unsigned integer in octal notation.
- `s`: null terminated string.
- `a, A`: double value in hexadecimal notation


### Example: basic `sprintf()`

Let's begin with a minimal example to explore the different formatting options
of `sprintf()`. Consider a real fraction like `1/6`; in R the default output 
of this fraction will be:

```{r}
1 / 6
```

Notice that `1/6` is printed with seven decimal digits. The number `1/6` is 
actually an irrational number and so the computer needs to round it to some 
number of decimal digits. You can modify the default printing format in several 
ways. One option is to display only six decimal digits with the `%f` option:

```{r}
# print 6 decimals
sprintf('%f', 1/6)
```

But you can also specify a different number of decimal digits, say 3. This can 
be achieved specifying an option of `%.3f`:

```{r}
# print 3 decimals
sprintf('%.3f', 1/6)
```

The table below shows six different outputs for `1/6`

| Notation | Output                   |
|:---------|:-------------------------|
| `%s`    | `r sprintf('%s', 1/6)`   |
| `%f`    | `r sprintf('%f', 1/6)`   |
| `%.3f`  | `r sprintf('%.3f', 1/6)` |
| `%e`    | `r sprintf('%e', 1/6)`   |
| `%E`    | `r sprintf('%E', 1/6)`   |
| `%g`    | `r sprintf('%g', 1/6)`   |


When would you use `sprintf()`? Everytime you produce output text. Some 
cases include: 

1) exporting output to some file.
2) printing output on console.
3) forming new strings.


### Example: File Names

When working on data analysis projects, it is common to generate different 
files with similar names (e.g. either for creating images, or data files, or 
documents). Imagine that you need to generate the names of 3 data files 
(with .csv extension). All the files have the same prefix name but each of them
has a different number: `data01.csv`, `data02.csv`, and `data03.csv`. 
One naive solution to generate a character vector with these names in R would 
be to write something like this:

```{r file_names1}
file_names <- c('data01.csv', 'data02.csv', 'data03.csv')
```

Instead of writing each file name, you can generate the vector `file_names`
in a more efficient way taking advantage of the vectorized nature of 
`paste0()`:

```{r file_names2}
file_names <- paste0('data0', 1:3, '.csv')

file_names
```

Now imagine that you need to generate 100 file names numbered from 01, 02, 03, 
to 100. You could write a vector with 100 file names but it's going to take 
you a while. A preferable solution is to use `paste0()` like in the approach 
of the previous example. In this case however, you would need to create two 
separate vectors---one with numbers 01 to 09, and another one with numbers 
10 to 100---and then concatenate them in one single vector:

```{r file_names3}
files1 <- paste0('data0', 1:9, '.csv')
files2 <- paste0('data', 10:100, '.csv')
file_names <- c(files1, files2)
```

Instead of using `paste0()` to create two vectors, you can use `sprintf()` 
with the `%0xd` option to indicate that an integer should be padded with `x` 
leading zeros. For instance, the first nine file names can be generated as:

```{r file_names4}
sprintf('data%02d.csv', 1:9)
```

To generate the 100 file names do:

```{r file_names5}
file_names <- sprintf('data%02d.csv', 1:100)
```

The first nine elements in `file_names` will include a leading zero before 
the integer; the following elements will not include the leading zero.


### Example: Fahrenheit to Celsius

This example involves working on a function to convert Fahrenheit degrees
into Celsius degrees. The conversion formula is:

$$
Celsius = (Fahrenheit - 32) \times \frac{5}{9}
$$

You can define a simple function `to_celsius()` that takes one argument,
`temp`, which is a number representing temperature in Fahrenheit degrees. 
This function will return the temperature in Celsius degrees:

```{r to_celsius}
to_celsius <- function(temp = 1) {
  (temp - 32) * 5/9
}
```

You can use `to_celsius()` as any other function in R. Say you want to 
know how many Celsius degrees are 95 Fahrenheit degrees:

```{r to_celsius1}
to_celsius(95)
```

To make things more interesting, let's create another function that not only 
computes the temperature conversion but also prints a more informative message, something like: `95 Fahrenheit degrees = 35 Celsius degrees`. 

We'll name this function `fahrenheit2celsius()`:

```{r fahrenheit2celsius}
fahrenheit2celsius <- function(temp = 1) {
  celsius <- to_celsius(temp)
  sprintf('%.2f Fahrenheit degrees = %.2f Celsius degrees', temp, celsius)
}
```

Notice that `fahrenheit2celsius()` makes use of `to_celsius()` to
compute the Celsius degrees. And then `sprintf()` is used with the options
`%.2f` to display the temperatures with two decimal digits. Try it out:

```{r fah2celsius}
fahrenheit2celsius(95)
fahrenheit2celsius(50)
```


### Example: Car Traveled Distance

Our third example is a little bit more sophisticated. The idea is to construct
an object of class `"car"` that contains characteristics like the name 
of the car, its make, its year, and its fuel consumption in _city_, 
_highway_ and _combined_.

Let's consider a _Mazda 3_ for this example. One possible way to define
a `"car"` object is to use a list with the following elements:

```{r mazda3}
mazda3 <- list(
  name = 'mazda3', # car name
  make = 'mazda',  # car make
  year = 2015,     # year model
  city = 30,       # fuel consumption in city
  highway = 40,    # fuel consumption in highway
  combined = 33)   # fuel consumption combined (city-and-hwy)
```

So far we have an object `mazda3` that is essentially a list. Because we
want to create a `print()` method for objects of class `"car"` we
need to assign this class to our `mazda3`:

```{r mazda3_car}
class(mazda3) <- "car"
```

Now that we have our `"car"` object, we can create a `print.car()` function. 
In this way, everytime we type `mazda3`, instead of getting the typical list 
output, we will get a customized display:

```{r print_car}
print.car <- function(x) {
  cat("Car\n")
  cat(sprintf('name: %s\n', x$name))
  cat(sprintf('make: %s\n', x$make))
  cat(sprintf('year: %s\n', x$year))
  invisible(x)
}
```

Next time you type `mazda3` in your console, R will display these lines:

```{r print_mazda3}
mazda3
```

It would be nice to have a function `miles()` that allows you to calculate the 
traveled distance for a given amount of gas (in gallons), taking into account 
the type of fuel consumption (e.g. city, highway, combined):

```{r miles_function}
miles <- function(car, fuel = 1, mpg = 'city') {
  stopifnot(class(car) == 'car')
  switch(mpg,
         'city' = car$city * fuel,
         'highway' = car$highway * fuel,
         'combined' = car$combined * fuel,
         car$city * fuel)
}
```


The `miles()` function takes three parameters: `car` is an object of
class `"car"`, `fuel` is the number of gallons, and `mpg` is the
type of fuel consumption (`'city', `highway`, `combined'`). The first
command checks whether the first parameter is an object of class `"car"`.
If it is not, then the function will stop the execution raising an error. 
The second command involves using the function `switch()` to compute the 
traveled miles. It switches to the corresponding consumption depending on the
provided value of `mpg`. Note that the very last switch condition is a
_safety_ condition in case the user mispecifies `mpg`.

Let's say you want to know how many miles the `mazda3` could
travel with 4 gallons of gas depending on the different types of consumption:

```{r miles_test}
miles(mazda3, fuel = 4, 'city')
miles(mazda3, fuel = 4, 'highway')
miles(mazda3, fuel = 4, 'combined')
```

Again, to make things more user friendly, we are going to create a function
`get_distance()` that prints a more informative message about the
traveled distance:

```{r get_distance_function}
get_distance <- function(car, fuel = 1, mpg = 'city') {
  distance <- miles(car, fuel = fuel, mpg = mpg) 
  cat(sprintf('A %s can travel %s miles\n',
              car$name, distance))
  cat(sprintf('with %s gallons of gas\n', fuel))
  cat(sprintf('using %s consumption', mpg))
}
```

And this is how the output when calling `get_distance` looks like:

```{r get_distance_test}
get_distance(mazda3, 4, 'city')
```


### Example: Coffee Prices

Consider some coffee drinks and their prices. We'll put this information in a 
vector like this:

```{r coffee_prices}
prices <- c(
  'americano' = 2, 
  'latte' = 2.75, 
  'mocha' = 3.45, 
  'capuccino' = 3.25)
```

What type of vector is `prices`? Is it a character vector? Is it numeric 
vector? Or is it some sort of vector with mix-data? We have seen that vectors 
are _atomic_ structures, meaning that all their elements must be of the
same class. So `prices` is definitely not a vector with mix-data. 
From the code chunk we can observe that each element of the vector is formed 
by a string, followed by the `=` sign, followed by some number. 
This way of defining a vector is not very common in R but it is perfectly 
valid. Each string represents the name of an element, while 
the numbers are the actual elements. Therefore `prices` is in reality 
a numeric vector. You can confirm this by looking at the mode (or data type):

```{r mode_prices}
mode(prices)
```

Let's say you want to list the names of the coffees and their prices. If you 
just simply try to `print()` the prices, the output will be the entire 
vector `prices`:

```{r print_prices}
print(prices)
```

Alternatively, you can use a for loop to `print()` each individual element of 
the vector `prices`, but again the output is displayed in an awkward fashion:

```{r for_loop_print_prices}
for (p in seq_along(prices)) {
  print(prices[p])
}
```

To list the names of the coffees and their prices, it would be nicer to use
a combination of `paste0()` and `print()`. In addition, you can be
more descriptive adding some auxiliary text such that the output prints
something like: _"americano has a price of $2"_.

```{r paste_prices}
for (p in seq_along(prices)) {
  print(paste0(names(prices)[p], ' has price of $', prices[p]))
}
```

Another possible solution consists of combining `print()` and `sprintf()`:

```{r sprintf_prices}
for (p in seq_along(prices)) {
  print(sprintf('%s has price of $%s', names(prices)[p], prices[p]))
}
```

One limitation of `quote()` is that it won't work inside a for loop:

```{r noquote_prices}
for (p in seq_along(prices)) {
  noquote(sprintf('%s has price of $%s', names(prices)[p], prices[p]))
}
```

If what you want is to print the output wihtout quotes, then you need to 
use `cat()`; just make sure to add a newline character `"\n"`:

```{r cat_prices}
for (p in seq_along(prices)) {
  cat(sprintf('%s has price of $%s\n', names(prices)[p], prices[p]))
}
```

