# Stringr Basics


## Introduction

So far we've seen the various functions R provides to perform basic string 
processing and manipulations of `"character"` data. Most of the times these 
functions are enough and they will allow you to get your job done. However, 
they have some drawbacks. For instance, consider the following example:

```{r ugly_nchar}
# some text vector
text <- c("one", "two", "three", NA, "five")

# how many characters in each string?
nchar(text)
```

As you can tell, `nchar()` gives `NA` a value of 2, as if it were a string 
formed by two characters. Perhaps this may be acceptable in some cases, but 
taking into account all the operations in R, it would be better to leave `NA` 
as a missing value, instead of treating it as a string of two characters.

Another awkward example can be found with `paste()`. The default separator is 
a blank space, which more often than not is what you want to use. But that's 
secondary. The really annoying thing is when you want to paste things that 
include zero length arguments (e.g. `NULL`, `character(0)`). How does `paste()` 
behave in those cases? See below:  

```{r paste_berkeley_ex, tidy=FALSE}
# this works fine
paste("University", "of", "California", "Berkeley")

# this works fine too
paste("University", "of", "California", "Berkeley")

# this is weird
paste("University", "of", "California", "Berkeley", NULL)

# this is ugly
paste("University", "of", "California", "Berkeley", NULL, character(0), 
      "Go Bears!")
```

Notice the output from the last example (the _ugly_ one). The objects `NULL` 
and `character(0)` have zero length, yet when included inside `paste()` they 
are treated as an empty string `""`. Wouldn't be good if `paste()` removed 
zero length arguments? Sadly, there's nothing we can do to change `nchar()` and 
`paste()`. But fear not. There is a very nice package that solves these 
problems and provides several functions for carrying out consistent string 
processing.



## Package `stringr`

Thanks to Hadley Wickham, we have the package `stringr` that adds more 
functionality to the base functions for handling strings in R. 

[http://cran.r-project.org/web/packages/stringr/index.html](http://cran.r-project.org/web/packages/stringr/index.html)

According to the description of the package:

> "is a set of simple wrappers that make R's string functions more consistent, 
> simpler and easier to use. It does this by ensuring that: function and 
> argument names (and positions) are consistent, all functions deal with NA's 
> and zero length character appropriately, and the output data structures from 
> each function matches the input data structures of other functions."

To install `"stringr"` use the function `install.packages()`. Once installed, 
load it to your current session with `library()`:

```{r install_stringr, message=FALSE, eval=FALSE}
# installing 'stringr'
install.packages("stringr")

# load 'stringr'
library(stringr)
```

```{r load_stringr, echo=FALSE}
library(stringr)
```


## Basic String Operations

`"stringr"` provides functions for both 1) basic manipulations and 2) for 
regular expression operations. In this chapter we cover those functions that 
have to do with basic manipulations. 

The following table contains the `stringr` functions for basic string operations:

| Function       | Description                             | Similar to    |
|:---------------|:----------------------------------------|:--------------|
| `str_c()`      | string concatenation                    | `paste()`     |
| `str_length()` | number of characters                    | `nchar()`     |
| `str_sub()`    | extracts substrings                     | `substring()` |
| `str_dup()`    | duplicates characters                   | _none_        |
| `str_trim()`   | removes leading and trailing whitespace | _none_        |
| `str_pad()`    | pads a string                           | _none_        |
| `str_wrap()`   | wraps a string paragraph                | `strwrap()`   |
| `str_trim()`   | trims a string                          | _none_        |


Notice that all functions in `stringr` start with `"str_"` followed by a term 
associated to the task they perform. For example, `str_length()` gives you the 
number (i.e. length) of characters in a string. In addition, some functions are 
designed to provide a better alternative to already existing functions. This is 
the case of `str_length()` which is intended to be a substitute of `nchar()`. 
Other functions, however, don't have a corresponding alternative such as 
`str_dup()` which allows you to duplicate characters.


### Concatenating with `str_c()`

Let's begin with `str_c()`. This function is equivalent to `paste()` but 
instead of using the white space as the default separator, `str_c()` uses the 
empty string `""` which is a more common separator when _pasting_ strings:

```{r str_c_ex}
# default usage
str_c("May", "The", "Force", "Be", "With", "You")

# removing zero length objects
str_c("May", "The", "Force", NULL, "Be", "With", "You", character(0))
```

Observe another major difference between `str_c()` and `paste()`: zero length 
arguments like `NULL` and `character(0)` are silently removed by `str_c()`. 

If you want to change the default separator, you can do that as usual by 
specifying the argument `sep`:

```{r str_join}
# changing separator
str_c("May", "The", "Force", "Be", "With", "You", sep = "_")
```



### Number of characters with `str_length()`

As we've mentioned before, the function `str_length()` is equivalent to 
`nchar()`. Both functions return the number of characters in a string, that is, 
the _length_ of a string (do not confuse it with the `length()` of a vector). 
Compared to `nchar()`, `str_length()` has a more consistent behavior when 
dealing with `NA` values. Instead of giving `NA` a length of 2, `str_length()` 
preserves missing values just as `NA`s.

```{r str_length_ex}
# some text (NA included)
some_text <- c("one", "two", "three", NA, "five")

# compare 'str_length' with 'nchar'
nchar(some_text)
str_length(some_text)
```

In addition, `str_length()` has the nice feature that it converts factors to 
characters, something that `nchar()` is not able to handle:

```{r str_length_with_factors}
some_factor <- factor(c(1,1,1,2,2,2), labels = c("good", "bad"))
some_factor

# try 'nchar' on a factor
nchar(some_factor)

# now compare it with 'str_length'
str_length(some_factor)
```


### Substring with `str_sub()`

To extract substrings from a character vector `stringr` provides `str_sub()` 
which is equivalent to `substring()`. The function `str_sub()` has the 
following usage form:

```
str_sub(string, start = 1L, end = -1L)
```

The three arguments in the function are: a `string` vector, a `start` value 
indicating the position of the first character in substring, and an `end` value 
indicating the position of the last character. Here's a simple example with a 
single string in which characters from 1 to 5 are extracted:

```{r str_sub_ex1}
lorem <- "Lorem Ipsum"

# apply 'str_sub'
str_sub(lorem, start = 1, end = 5)

# equivalent to 'substring'
substring(lorem, first = 1, last = 5)

# another example
str_sub("adios", 1:3)
```

An interesting feature of `str_sub()` is its ability to work with negative 
indices in the `start` and `end` positions. When we use a negative position, `str_sub()` counts backwards from last character: 

```{r str_sub_ex2}
resto = c("brasserie", "bistrot", "creperie", "bouchon")

# 'str_sub' with negative positions
str_sub(resto, start = -4, end = -1)

# compared to substring (useless)
substring(resto, first = -4, last = -1)
```

Similar to `substring()`, we can also give `str_sub()` a set of positions which 
will be recycled over the string. But even better, we can give `str_sub()` 
a negative sequence, something that `substring()` ignores:

```{r str_sub_ex3}
# extracting sequentially
str_sub(lorem, seq_len(nchar(lorem)))
substring(lorem, seq_len(nchar(lorem)))

# reverse substrings with negative positions
str_sub(lorem, -seq_len(nchar(lorem)))
substring(lorem, -seq_len(nchar(lorem)))
```

We can use `str_sub()` not only for extracting subtrings but also for replacing
substrings:

```{r str_sub_ex4}
# replacing 'Lorem' with 'Nullam'
lorem <- "Lorem Ipsum"
str_sub(lorem, 1, 5) <- "Nullam"
lorem

# replacing with negative positions
lorem <- "Lorem Ipsum"
str_sub(lorem, -1) <- "Nullam"
lorem

# multiple replacements 
lorem <- "Lorem Ipsum"
str_sub(lorem, c(1,7), c(5,8)) <- c("Nullam", "Enim")
lorem
```


### Duplication with `str_dup()`

A common operation when handling characters is _duplication_. The problem is 
that R doesn't have a specific function for that purpose. But `stringr` does: `str_dup()` duplicates and concatenates strings within a character vector. 
Its usage requires two arguments:

```
str_dup(string, times)
```

The first input is the `string` that you want to dplicate. The second input, 
`times`, is the number of times to duplicate each string:

```{r str_dup_ex}
# default usage
str_dup("hola", 3)

# use with differetn 'times'
str_dup("adios", 1:3)

# use with a string vector
words <- c("lorem", "ipsum", "dolor", "sit", "amet")
str_dup(words, 2)

str_dup(words, 1:5)
```


### Padding with `str_pad()`

Another handy function that we can find in `stringr` is `str_pad()` for 
_padding_ a string. Its default usage has the following form:

```
str_pad(string, width, side = "left", pad = " ")
```

The idea of `str_pad()` is to take a string and pad it with leading or trailing 
characters to a specified total `width`. The default padding character is a 
space (`pad = " "`), and consequently the returned string will appear to be 
either left-aligned (`side = "left"`), right-aligned (`side = "right"`), or 
both (`side = "both"`). 

Let's see some examples:

```{r str_pad_ex}
# default usage
str_pad("hola", width = 7)

# pad both sides
str_pad("adios", width = 7, side = "both")

# left padding with '#'
str_pad("hashtag", width = 8, pad = "#")

# pad both sides with '-'
str_pad("hashtag", width = 9, side = "both", pad = "-")
```


### Wrapping with `str_wrap()`

The function `str_wrap()` is equivalent to `strwrap()` which can be used to 
_wrap_ a string to format paragraphs. The idea of wrapping a (long) string is 
to first split it into paragraphs according to the given `width`, and then add 
the specified indentation in each line (first line with `indent`, following 
lines with `exdent`). Its default usage has the following form:

```
str_wrap(string, width = 80, indent = 0, exdent = 0)
```

For instance, consider the following quote (from Douglas Adams) converted into 
a  paragraph:

```{r douglas_adams, tidy=FALSE}
# quote (by Douglas Adams)
some_quote <- c(
  "I may not have gone",
  "where I intended to go,", 
  "but I think I have ended up",
  "where I needed to be")

# some_quote in a single paragraph
some_quote <- paste(some_quote, collapse = " ")
```

Now, say you want to display the text of `some_quote` within some pre-specified 
column width (e.g. width of 30). You can achieve this by applying `str_wrap()` 
and setting the argument `width = 30`

```{r str_wrap_ex1, tidy=FALSE}
# display paragraph with width=30
cat(str_wrap(some_quote, width = 30))
```

Besides displaying a (long) paragraph into several lines, you may also wish to 
add some indentation. Here's how you can indent the first line, as well as the 
following lines: 

```{r str_wrap_ex2, tidy=FALSE}
# display paragraph with first line indentation of 2
cat(str_wrap(some_quote, width = 30, indent = 2), "\n")

# display paragraph with following lines indentation of 3
cat(str_wrap(some_quote, width = 30, exdent = 3), "\n")
```



### Trimming with `str_trim()`

One of the typical tasks of string processing is that of parsing a text into 
individual words. Usually, you end up with words that have blank spaces, called _whitespaces_, on either end of the word. In this situation, you can use the 
`str_trim()` function to remove any number of whitespaces at the ends of a 
string. Its usage requires only two arguments:

```
str_trim(string, side = "both")
```

The first input is the `string` to be strimmed, and the second input indicates 
the `side` on which the whitespace will be removed.

Consider the following vector of strings, some of which have whitespaces either 
on the left, on the right, or on both sides. Here's what `str_trim()` would do 
to them under different settings of `side`

```{r steal, tidy=FALSE}
# text with whitespaces
bad_text <- c("This", " example ", "has several   ", "   whitespaces ")

# remove whitespaces on the left side
str_trim(bad_text, side = "left")

# remove whitespaces on the right side
str_trim(bad_text, side = "right")

# remove whitespaces on both sides
str_trim(bad_text, side = "both")
```


### Word extraction with `word()`

We end this chapter describing the `word()` function that is designed to 
extract words from a sentence:

```
word(string, start = 1L, end = start, sep = fixed(" "))
```

The way in which you use `word()` is by passing it a `string`, together with a 
`start` position of the first word to extract, and an `end` position of the 
last word to extract. By default, the separator `sep` used between words is a 
single space.

Let's see some examples:

```{r word_ex}
# some sentence
change <- c("Be the change", "you want to be")

# extract first word
word(change, 1)

# extract second word
word(change, 2)

# extract last word
word(change, -1)

# extract all but the first words
word(change, 2, -1)
```

`stringr` has more functions but we'll discuss them in the chapters about 
[regular expressions](#regex1).
